---
title: "Client comparison in Medalla"
author:
- name: Barnabé Monnot
  url: https://twitter.com/barnabemonnot
  affiliation: Robust Incentives Group, Ethereum Foundation
  affiliation_url: https://github.com/ethereum/rig
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
description: |
  Looking at statistics per (declared) client.
---

```{r setup, include=FALSE}
library(tidyverse)
library(glue)
library(RPostgres)
library(DBI)
library(ineq)
library(PKI)
library(rmarkdown)
library(zoo)
library(data.table)

source(here::here("notebooks/lib.R"))
source(here::here("notebooks/pw.R"))

con <- dbConnect(RPostgres::Postgres(), user="chain", password=pw)

options(digits=10)
options(scipen = 999) 

# Make the plots a bit less pixellated
knitr::opts_chunk$set(dpi = 400)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# A minimal theme I like
newtheme <- theme_grey() + theme(
  axis.text = element_text(size = 9),
  axis.title = element_text(size = 12),
  axis.line = element_line(colour = "#000000"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  legend.title = element_text(size = 12),
  legend.text = element_text(size = 10),
  legend.box.background = element_blank(),
  legend.key = element_blank(),
  strip.text.x = element_text(size = 10),
  strip.background = element_rect(fill = "white")
)
theme_set(newtheme)

myred <- "#F05431"
myyellow <- "#FED152"
mygreen <- "#BFCE80"
```

```{r}
until_epoch <- 12125
slot_chunk_res <- 25000
epoch_resolution <- 12125

slots_per_epoch <- 32
until_slot <- until_epoch * slots_per_epoch
slots_per_year <- 365.25 * 24 * 60 * 60 / 12
epochs_per_year <- slots_per_year / slots_per_epoch
```

```{r}
all_bxs <- fread(here::here("rds_data/all_bxs.csv"))
```

```{r}
all_ats <- fread(here::here("rds_data/all_ats.csv"))
```

```{r}
blocks_per_client <- all_bxs %>%
  .[declared_client!="undecided" & slot < 30000, .(slot, declared_client)] %>%
  merge(fread(here::here("rds_data/subset_ats_30000.csv")) %>%
          .[,.(n_subset=sum(n_subset)),by=slot],
        all.x = TRUE) %>%
  setnafill("const", 0, cols="n_subset") %>%
  mutate(`Collection` = "0 - 30000") %>%
  union(
    all_bxs %>%
      .[declared_client!="undecided" & slot >= 590000 & slot < 610000, .(slot, declared_client)] %>%
      merge(fread(here::here("rds_data/subset_ats_590000+.csv")) %>%
              .[,.(n_subset=sum(n_subset)),by=slot],
            all.x = TRUE) %>%
      setnafill("const", 0, cols="n_subset") %>%
      mutate(`Collection` = "590000 - 610000")
  )
```

Declared client is obtained in the graffiti:

- Either when the graffiti starts with `poap` and ends with `a`, `b`, `c`, `d` and `e` (respectively, Prysm, Lighthouse, Teku, Nimbus and Lodestar)
- Or when the graffiti contains the client name in its string (e.g., `teku/v0.12.4-dev-4855f89c`)

## Aggregates

### Aggregates included per block

```{r}
all_ats %>%
  .[slot < 3e4 | (slot >= 59e4 & slot < 61e4), .(included_ats = .N), by=slot] %>%
  merge(blocks_per_client) %>%
  filter(declared_client != "lodestar") %>%
  ggplot(aes(x = declared_client, y = included_ats, fill = Collection)) +
  stat_summary(geom = "bar", fun = mean, position = "dodge") +
  stat_summary(geom = "errorbar", fun.data = mean_se, position = position_dodge(.9), width = 0.2) +
  scale_fill_manual(values = c(myred, myyellow)) +
  ggtitle("Average number of aggregates included per block") +
  xlab("Declared client") +
  ylab("Average number of aggregates")
```

### Inclusion of subset aggregates

[Subset aggregates](https://ethereum.github.io/rig/medalla-data-challenge/notebooks/explore.html) are aggregates included in a block which are fully covered by another aggregate included in the same block. Namely, when aggregate 1 has attesting indices $I$ and aggregate 2 has attesting indices $J$, aggregate 1 is a subset aggregate when $I \subset J$.

```{r}
blocks_per_client %>%
  filter(declared_client != "lodestar") %>%
  ggplot(aes(x = declared_client, y = n_subset, fill = Collection)) +
  stat_summary(geom = "bar", fun = mean, position = "dodge") +
  stat_summary(geom = "errorbar", fun.data = mean_se, position = position_dodge(.9), width = 0.2) +
  scale_fill_manual(values = c(myred, myyellow)) +
  ggtitle("Average number of subset aggregates per block") +
  xlab("Declared client") +
  ylab("Average subset aggregates")
```
```{r}
all_ats %>%
  .[slot < 3e4 | (slot >= 59e4 & slot < 61e4), .(included_ats = .N), by=slot] %>%
  merge(blocks_per_client) %>%
  filter(declared_client != "lodestar") %>%
  group_by(declared_client, Collection) %>%
  summarise(percent = sum(n_subset) / sum(included_ats) * 100) %>%
  ggplot(aes(x = declared_client, y = percent, fill = Collection)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(myred, myyellow)) +
  ggtitle("Percentage of subset aggregates among included aggregates") +
  xlab("Declared client") +
  ylab("Percentage of subset aggregates in block")
```



