---
title: "Pyrmont client comparison"
author:
- name: Barnabé Monnot
  url: https://twitter.com/barnabemonnot
  affiliation: Robust Incentives Group, Ethereum Foundation
  affiliation_url: https://github.com/ethereum/rig
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
description: |
  Onwards!
---

```{r setup, include=FALSE}
library(tidyverse)
library(data.table)
library(rmarkdown)
library(infer)

source(here::here("notebooks/lib.R"))

options(digits=10)
options(scipen = 999) 

# Make the plots a bit less pixellated
knitr::opts_chunk$set(dpi = 300)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# A minimal theme I like
newtheme <- theme_grey() + theme(
  axis.text = element_text(size = 9),
  axis.title = element_text(size = 12),
  axis.line = element_line(colour = "#000000"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  legend.title = element_text(size = 12),
  legend.text = element_text(size = 10),
  legend.box.background = element_blank(),
  legend.key = element_blank(),
  strip.text.x = element_text(size = 10),
  strip.background = element_rect(fill = "white")
)
theme_set(newtheme)

myred <- "#F05431"
myyellow <- "#FED152"
mygreen <- "#BFCE80"

end_epoch <- 1020
slots_per_epoch <- 32
until_slot <- (end_epoch + 2) * slots_per_epoch - 1
slot_chunk_res <- until_slot %/% 15
slots_per_year <- 365.25 * 24 * 60 * 60 / 12
epochs_per_year <- slots_per_year / slots_per_epoch
```

```{r}
validators <- fread(here::here("pyrmont_data/initial_validators.csv"))
```

## Reward rates

```{r}
start_epoch <- 675
end_epoch <- 900

start_balances <- get_balances_active_validators(start_epoch)[
    validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
  ] %>%
  mutate(
    measurement_epoch = start_epoch
  ) %>%
  select(-time_active, -activation_epoch)

end_balances <- get_balances_active_validators(end_epoch)[
    validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
  ] %>%
  mutate(
    measurement_epoch = end_epoch
  ) %>%
  select(-time_active, -activation_epoch)

reward_rates <- start_balances %>%
  inner_join(end_balances,
             by = c("validator_index", "client", "first_digit")) %>%
  mutate(reward_rate = (balance.y - balance.x) / balance.x * 100 * epochs_per_year / (measurement_epoch.y - measurement_epoch.x))
```

```{r}
# Differences by client
mean_diff <- reward_rates %>% 
  filter(client == "lighthouse" | client == "prysm") %>%
  specify(formula = reward_rate ~ client) %>% 
  calculate(stat = "diff in means", order = c("lighthouse", "prysm"))

null_distribution <- reward_rates %>% 
  filter(client == "lighthouse" | client == "prysm") %>%
  specify(formula = reward_rate ~ client) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("lighthouse", "prysm"))

# null_distribution %>% 
#   get_pvalue(obs_stat = mean_diff, direction = "both")
```

```{r}
# Differences by node region
mean_diff <- reward_rates %>%
  mutate(node_group = if_else(first_digit <= 2, "A", "B")) %>%
  specify(formula = reward_rate ~ node_group) %>% 
  calculate(stat = "diff in means", order = c("A", "B"))

null_distribution <- reward_rates %>%
  mutate(node_group = if_else(first_digit <= 2, "A", "B")) %>%
  specify(formula = reward_rate ~ node_group) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("A", "B"))

# null_distribution %>% 
#   get_pvalue(obs_stat = mean_diff, direction = "both")
```

```{r}
# pairwise.t.test(reward_rates$reward_rate, reward_rates$client, p.adjust.method = "none")
```

```{r}
# reward_rates %>%
#   group_by(client) %>%
#   summarise(reward_rate = mean(reward_rate))
```

```{r}
# reward_rates %>%
#   mutate(node_group = if_else(first_digit <= 2, "A", "B")) %>%
#   group_by(node_group) %>%
#   summarise(reward_rate = mean(reward_rate))
```

```{r}
# pairwise.t.test(reward_rates$reward_rate, reward_rates$first_digit, p.adjust.method = "none")
```

```{r message=FALSE}
get_reward_timelines <- function(start_epoch, end_epoch, step=25) {
  start_balances <- get_balances_active_validators(start_epoch)[
    validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
  ] %>%
    mutate(
      measurement_epoch = start_epoch
    ) %>%
    select(-time_active, -activation_epoch)
  
  seq(start_epoch+step, end_epoch+1, step) %>%
    map(function(epoch) {
      end_balances <- get_balances_active_validators(epoch)[
        validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
      ] %>%
        mutate(
          measurement_epoch = epoch
        ) %>%
        select(-time_active, -activation_epoch)
      
      t <- start_balances %>%
        inner_join(end_balances,
                   by = c("validator_index", "client", "first_digit")) %>%
        mutate(reward_rate = (balance.y - balance.x) / balance.x * 100 * epochs_per_year / (measurement_epoch.y - measurement_epoch.x))
      rr <- t %>%
        group_by(group = client, measurement_epoch.y) %>%
        summarise(avg_rr = mean(reward_rate), n_group = n()) %>%
        mutate(name = "client") %>%
        union(
          t %>%
            group_by(group = as.character(first_digit), measurement_epoch.y) %>%
            summarise(avg_rr = mean(reward_rate), n_group = n()) %>%
            mutate(name = "region")
        )
      
      start_balances <- end_balances
      return(rr)
    }) %>%
    bind_rows()
}
```

### Since genesis

```{r message=FALSE}
rr_series <- get_reward_timelines(1, 1090, step=50)
```

We first look at the reward rates per client since genesis.

```{r}
rr_series %>%
  filter(name == "client") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per client") +
  scale_color_discrete(name = "Client")
```

Clients are hosted on AWS nodes scattered across four regions in roughly equal proportions. We look at the reward rates per region.

```{r}
rr_series %>%
  filter(name == "region") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per region") +
  scale_color_discrete(name = "Region")
```

Performing an omnibus test to detect significant difference between any of the four groups, we are unable to find such significance.

### A natural experiment: scaling down nodes

Around epoch 1020, nodes from regions 1 and 2 were scaled down from t3.xlarge units (4cpu 16GB mem, with unlimited cpu burst) to m5.large units (2cpu, 8GB mem, no burst). We observe a significant loss of performance despite continuous uptime.

```{r message=FALSE}
rr_series <- get_reward_timelines(900, 1090, step=10)
```

```{r}
rr_series %>%
  filter(name == "region") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per region") +
  scale_color_discrete(name = "Region")
```

Reward rates per client are affected in roughly equal proportions.

```{r}
rr_series %>%
  filter(name == "client") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per client") +
  scale_color_discrete(name = "Client")
```

