---
title: "Pyrmont data"
author:
- name: Barnabé Monnot
  url: https://twitter.com/barnabemonnot
  affiliation: Robust Incentives Group, Ethereum Foundation
  affiliation_url: https://github.com/ethereum/rig
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
description: |
  Onwards!
---

```{r setup, include=FALSE}
library(tidyverse)
library(data.table)
library(patchwork)
library(rmarkdown)

source(here::here("notebooks/lib.R"))

options(digits=10)
options(scipen = 999) 

# Make the plots a bit less pixellated
knitr::opts_chunk$set(dpi = 300)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# A minimal theme I like
newtheme <- theme_grey() + theme(
  axis.text = element_text(size = 9),
  axis.title = element_text(size = 12),
  axis.line = element_line(colour = "#000000"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  legend.title = element_text(size = 12),
  legend.text = element_text(size = 10),
  legend.box.background = element_blank(),
  legend.key = element_blank(),
  strip.text.x = element_text(size = 10),
  strip.background = element_rect(fill = "white")
)
theme_set(newtheme)

myred <- "#F05431"
myyellow <- "#FED152"
mygreen <- "#BFCE80"

slots_per_epoch <- 32
```

```{r eval=FALSE}
blx_and_ats <- 0:220 %>%
  map(get_blocks_and_attestations) %>%
  purrr::transpose() %>%
  map(rbindlist)
```

```{r eval=FALSE}
all_bxs <- copy(blx_and_ats$block)
all_bxs[, declared_client := find_client(graffiti)]

all_ats <- copy(blx_and_ats$attestations)
rm(blx_and_ats)
```

```{r eval=FALSE}
committees <- 0:220 %>%
  map(get_committees) %>%
  rbindlist()
```

```{r eval=FALSE}
all_bxs %>% fwrite(here::here("pyrmont_data/all_bxs.csv"))
all_ats %>% fwrite(here::here("pyrmont_data/all_ats.csv"))
committees %>% fwrite(here::here("pyrmont_data/committees.csv"))
```

```{r}
all_bxs <- fread(here::here("pyrmont_data/all_bxs.csv"))
all_ats <- fread(here::here("pyrmont_data/all_ats.csv"))
committees <- fread(here::here("pyrmont_data/committees.csv"))
```

```{r}
client_per_validator <- all_bxs %>%
  .[, .SD[which.max(slot)], by=proposer_index, .SDcols = c("slot", "declared_client")]
```

```{r}
plot_grid <- function(start_epoch, end_epoch, committees = NULL) {
  if (is.null(committees)) {
    committees <- start_epoch:end_epoch %>%
      map(get_committees) %>%
      rbindlist()
  }
  
  attestation_grid <- all_ats %>%
    .[att_slot >= start_epoch * slots_per_epoch & att_slot < (end_epoch + 1) * slots_per_epoch,
      .(att_slot, committee_index, attesting_indices)] %>%
    get_exploded_ats() %>%
    merge(committees,
          by.x = c("att_slot", "committee_index", "index_in_committee"),
          by.y = c("att_slot", "committee_index", "index_in_committee")) %>%
    .[,.(att_slot, epoch=att_slot %/% slots_per_epoch, validator_index, attested = 1)] %>%
    merge(client_per_validator %>%
            .[, .(proposer_index, declared_client)],
          by.x = c("validator_index"), by.y = c("proposer_index")) %>%
    unique()
  
  l = list()

  for (client in c("prysm", "lighthouse", "nimbus", "teku")) {
    attestation_grid_per_client <- attestation_grid %>%
      .[epoch <= end_epoch & epoch >= start_epoch & declared_client == client] %>%
      mutate(validator_index = as.factor(validator_index),
             attested = as_factor(attested))
    
    p <- attestation_grid_per_client %>%
      ggplot() +
      geom_tile(aes(x = epoch, y = validator_index, fill = attested)) +
      scale_fill_manual(values = c(mygreen), guide=FALSE) +
      scale_x_continuous(expand = c(0, 0)) +
      facet_wrap(vars(declared_client)) +
      theme(axis.text.y=element_blank(),
            axis.ticks.y=element_blank(),
            panel.background=element_rect(fill=myred, colour=myred),
            axis.title.x = element_text(size = 7),
            axis.title.y = element_text(size = 7),
            axis.text.x = element_text(size = 7),
            strip.text = element_text(size = 8))
    
    l[[client]] <- p
  }
  
  l[["prysm"]] | l[["lighthouse"]] | l[["nimbus"]] | l[["teku"]]
}
```

We look at data between epochs 0 and 220.

## Finding the declared client

We associate each validator index to a client among Lighthouse, Prysm, Nimbus, Teku and "undecided". To do so, for each canonical block produced, we inspect the graffiti. We associate the block to a client whenever the name of the client features in the graffiti (with a case-insensitive search). When the first four characters of the graffiti are `poap`, we check the last character: when the character is "a", "b", "c", "d" or "e", we guess the client producing the block is (respectively) Prysm, Lighthouse, Teku, Nimbus or Lodestar. Due to the low prevalence of Lodestar blocks, we have decided to remove it from the analysis.

We assume that a validator index is using the latest recorded client obtained from a block produced by that validator.

```{r}
client_per_validator %>%
  .[, .(count=.N), by=declared_client] %>%
  ggplot() +
  geom_col(aes(x = declared_client, y = count), fill = myred) +
  ggtitle("Distribution of clients in the dataset") +
  xlab("Declared client") +
  ylab("Count")
```

```{r}
start_epoch <- 0
end_epoch <- 99
```

### Epoch `r start_epoch` to `r end_epoch`

A moment in time, between `r get_date_from_epoch(start_epoch)` and `r get_date_from_epoch(end_epoch + 1)`.

```{r, layout="l-screen", fig.height=2}
plot_grid(start_epoch, end_epoch, committees)
```

```{r}
start_epoch <- 100
end_epoch <- 219
```

### Epoch `r start_epoch` to `r end_epoch`

A moment in time, between `r get_date_from_epoch(start_epoch)` and `r get_date_from_epoch(end_epoch + 1)`.

```{r, layout="l-screen", fig.height=2}
plot_grid(start_epoch, end_epoch, committees)
```