---
title: "Pyrmont client comparison"
author:
- name: Barnabé Monnot
  url: https://twitter.com/barnabemonnot
  affiliation: Robust Incentives Group, Ethereum Foundation
  affiliation_url: https://github.com/ethereum/rig
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
description: |
  Onwards!
---

```{r setup, include=FALSE}
library(tidyverse)
library(data.table)
library(rmarkdown)
library(infer)

source(here::here("notebooks/lib.R"))

options(digits=10)
options(scipen = 999) 

# Make the plots a bit less pixellated
knitr::opts_chunk$set(dpi = 300)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# A minimal theme I like
newtheme <- theme_grey() + theme(
  axis.text = element_text(size = 9),
  axis.title = element_text(size = 12),
  axis.line = element_line(colour = "#000000"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  legend.title = element_text(size = 12),
  legend.text = element_text(size = 10),
  legend.box.background = element_blank(),
  legend.key = element_blank(),
  strip.text.x = element_text(size = 10),
  strip.background = element_rect(fill = "white")
)
theme_set(newtheme)

myred <- "#F05431"
myyellow <- "#FED152"
mygreen <- "#BFCE80"

end_epoch <- 1020
slots_per_epoch <- 32
until_slot <- (end_epoch + 2) * slots_per_epoch - 1
slot_chunk_res <- until_slot %/% 15
slots_per_year <- 365.25 * 24 * 60 * 60 / 12
epochs_per_year <- slots_per_year / slots_per_epoch
```

```{r cache=TRUE}
all_ats <- fread(here::here("pyrmont_data/all_ats.csv"))
all_bxs <- fread(here::here("pyrmont_data/all_bxs.csv"))
block_root_at_slot <- get_block_root_at_slot(all_bxs)
validators <- fread(here::here("pyrmont_data/initial_validators.csv"))
```

```{r}
# start_epoch <- 675
# end_epoch <- 900
# 
# start_balances <- get_balances_active_validators(start_epoch)[
#     validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
#   ] %>%
#   mutate(
#     measurement_epoch = start_epoch
#   ) %>%
#   select(-time_active, -activation_epoch)
# 
# end_balances <- get_balances_active_validators(end_epoch)[
#     validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
#   ] %>%
#   mutate(
#     measurement_epoch = end_epoch
#   ) %>%
#   select(-time_active, -activation_epoch)
# 
# reward_rates <- start_balances %>%
#   inner_join(end_balances,
#              by = c("validator_index", "client", "first_digit")) %>%
#   mutate(reward_rate = (balance.y - balance.x) / balance.x * 100 * epochs_per_year / (measurement_epoch.y - measurement_epoch.x))
```

```{r}
# Differences by client
# mean_diff <- reward_rates %>% 
#   filter(client == "lighthouse" | client == "prysm") %>%
#   specify(formula = reward_rate ~ client) %>% 
#   calculate(stat = "diff in means", order = c("lighthouse", "prysm"))
# 
# null_distribution <- reward_rates %>% 
#   filter(client == "lighthouse" | client == "prysm") %>%
#   specify(formula = reward_rate ~ client) %>% 
#   hypothesize(null = "independence") %>% 
#   generate(reps = 1000, type = "permute") %>% 
#   calculate(stat = "diff in means", order = c("lighthouse", "prysm"))
# 
# null_distribution %>% 
#   get_pvalue(obs_stat = mean_diff, direction = "both")
```

```{r}
# Differences by node region
# mean_diff <- reward_rates %>%
#   mutate(node_group = if_else(first_digit <= 2, "A", "B")) %>%
#   specify(formula = reward_rate ~ node_group) %>% 
#   calculate(stat = "diff in means", order = c("A", "B"))
# 
# null_distribution <- reward_rates %>%
#   mutate(node_group = if_else(first_digit <= 2, "A", "B")) %>%
#   specify(formula = reward_rate ~ node_group) %>% 
#   hypothesize(null = "independence") %>% 
#   generate(reps = 1000, type = "permute") %>% 
#   calculate(stat = "diff in means", order = c("A", "B"))
# 
# null_distribution %>% 
#   get_pvalue(obs_stat = mean_diff, direction = "both")
```

```{r}
# pairwise.t.test(reward_rates$reward_rate, reward_rates$client, p.adjust.method = "none")
```

```{r}
# reward_rates %>%
#   group_by(client) %>%
#   summarise(reward_rate = mean(reward_rate))
```

```{r}
# reward_rates %>%
#   mutate(node_group = if_else(first_digit <= 2, "A", "B")) %>%
#   group_by(node_group) %>%
#   summarise(reward_rate = mean(reward_rate))
```

```{r}
# pairwise.t.test(reward_rates$reward_rate, reward_rates$first_digit, p.adjust.method = "none")
```

```{r message=FALSE}
get_reward_timelines <- function(start_epoch, end_epoch, step=25) {
  start_balances <- get_balances_active_validators(start_epoch)[
    validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
  ] %>%
    mutate(
      measurement_epoch = start_epoch
    ) %>%
    select(-time_active, -activation_epoch)
  
  seq(start_epoch+step, end_epoch+1, step) %>%
    map(function(epoch) {
      end_balances <- get_balances_active_validators(epoch)[
        validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
      ] %>%
        mutate(
          measurement_epoch = epoch
        ) %>%
        select(-time_active, -activation_epoch)
      
      t <- start_balances %>%
        inner_join(end_balances,
                   by = c("validator_index", "client", "first_digit")) %>%
        mutate(reward_rate = (balance.y - balance.x) / balance.x * 100 * epochs_per_year / (measurement_epoch.y - measurement_epoch.x))
      rr <- t %>%
        group_by(group = client, measurement_epoch.y) %>%
        summarise(avg_rr = mean(reward_rate), n_group = n()) %>%
        mutate(name = "client") %>%
        union(
          t %>%
            group_by(group = as.character(first_digit), measurement_epoch.y) %>%
            summarise(avg_rr = mean(reward_rate), n_group = n()) %>%
            mutate(name = "region")
        )
      
      start_balances <- end_balances
      return(rr)
    }) %>%
    bind_rows()
}
```

## Reward rates since genesis

```{r cache=TRUE, message=FALSE}
rr_series <- get_reward_timelines(1, 1101, step=50)
```

We first look at the reward rates per client since genesis.

```{r}
rr_series %>%
  filter(name == "client") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per client") +
  scale_color_discrete(name = "Client")
```

Clients are hosted on AWS nodes scattered across four regions in roughly equal proportions. We look at the reward rates per region.

```{r}
rr_series %>%
  filter(name == "region") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per region") +
  scale_color_discrete(name = "Region")
```

Performing an omnibus test to detect significant difference between any of the four groups, we are unable to find such significance.

## Experiment: Scaling down nodes

Around epoch 1020, nodes from regions 1 and 2 were scaled down from t3.xlarge units (4cpu 16GB mem, with unlimited cpu burst) to m5.large units (2cpu, 8GB mem, no burst). We observe a significant loss of performance despite continuous uptime.

```{r cache=TRUE, message=FALSE}
rr_series <- get_reward_timelines(900, 1100, step=10)
```

```{r}
rr_series %>%
  filter(name == "region") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per region") +
  scale_color_discrete(name = "Region")
```

Reward rates per client are affected in roughly equal proportions.

```{r}
rr_series %>%
  filter(name == "client") %>%
  ggplot(aes(x = measurement_epoch.y, y = avg_rr, group=group, color=group)) +
  geom_line() +
  xlab("Epoch") +
  ylab("Average reward rate") +
  ggtitle("Timeline of average rates of reward per client") +
  scale_color_discrete(name = "Client")
```

### Analysis by duty

```{r cache=TRUE}
start_epoch <- 950
end_epoch <- 1100
chunk_size <- 10

val_series <- get_stats_per_val(
  all_ats[att_slot >= start_epoch * slots_per_epoch & att_slot < end_epoch * slots_per_epoch],
  block_root_at_slot, chunk_size = chunk_size)
```

We look at four metrics across each region:

- Percentage of included attestations.
- Percentage of correct targets among expected attestations.
- Percentage of correct heads among expected attestations.
- Average inclusion delay.

To obtain a time series, we divide the period between epoch `r start_epoch` and epoch `r end_epoch` in chunks of size `r chunk_size` epochs. For each validator, we record how many included attestations appear in the dataset (ranging between 0 and `r chunk_size` for each chunk), the number of correct targets, correct heads and its average inclusion delay. We average over all validators in the EF-controlled set, measuring metrics either per client or per region.

We start by looking at the metrics per region.

```{r, layout="l-screen", fig.height=2}
val_series[
  validators[node_code > 0, .(validator_index, client, region=as.character(first_digit))], on="validator_index"
][
  , .(
    included_ats=sum(included_ats)/sum(expected_ats) * 100,
    correct_targets=sum(correct_targets)/sum(expected_ats) * 100,
    correct_heads=sum(correct_heads)/sum(expected_ats) * 100,
    inclusion_delay=mean(inclusion_delay, na.rm = TRUE)
  ),
  by=.(epoch, region)
] %>%
  melt(id.vars = c("epoch", "region")) %>%
  ggplot() +
  geom_line(aes(x = epoch, y = value, group=region, color=region)) +
  xlab("Epoch") +
  ylab("Value") +
  scale_color_discrete(name = "Region") +
  facet_wrap(vars(variable), nrow=1, scales="free_y") +
  theme(axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 8))
```

And now per client.

```{r, layout="l-screen", fig.height=2}
val_series[
  validators[node_code > 0, .(validator_index, client, first_digit)], on="validator_index"
][
  , .(
    included_ats=sum(included_ats)/sum(expected_ats) * 100,
    correct_targets=sum(correct_targets)/sum(expected_ats) * 100,
    correct_heads=sum(correct_heads)/sum(expected_ats) * 100,
    inclusion_delay=mean(inclusion_delay, na.rm = TRUE)
  ),
  by=.(epoch, client)
] %>%
  melt(id.vars = c("epoch", "client")) %>%
  ggplot() +
  geom_line(aes(x = epoch, y = value, group=client, color=client)) +
  xlab("Epoch") +
  ylab("Value") +
  scale_color_discrete(name = "Client") +
  facet_wrap(vars(variable), nrow=1, scales="free_y") +
  theme(axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 8))
```

